---
title: "01_data_preparation_CUP"
date: "`r Sys.Date()`"
engine: knitr
format: 
  html:
    self-contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    df-print: kable
---

## Introduction

This script processes a dataset where mutations and structural variants are written
in one column, split by comma's, to a new format where each gene has its own column.

This can later be used for an onkoprint.

## Preparation

```{r}
library(tidyverse)
library(here)
library(readxl)
```

Define functions

```{r}
df_wider_and_cleanup <- function(df) {
  df <- df |>
    mutate(col_name = str_replace_all(col_name, " ", "_")) |>
    pivot_wider(
      names_from = col_name,
      values_from = value_name,
      values_fill = NA,
      values_fn = ~ paste(.x, collapse = "|")
    )
  if (any(colnames(df) == "NA")) {
    df <- select(df, -"NA")
  }
  return(df)
}
```

## Load data

```{r}
file_location <- here()
input_excel <- "Oncoprin_CUPxMTB_syntheticdata_V3.2.xlsx"

# Already list the table of gene panels used. This is needed later to check whether genes were assessed
gene_panels_tsv <- "metadata_panels_and_sequencing_synthetic.tsv"
panel_per_platform_lookup_tsv <- "gene_panels_and_platform_names.tsv"
study_date <- Sys.Date()
cohort_name <- "CUP_synthetic"

CUP_data <- read_excel(file.path(file_location, "data", input_excel))
dim(CUP_data)
head(CUP_data, 1)
colnames(CUP_data)

(nr_of_samples <- nrow(CUP_data))
```

Study the data structure

```{r}
CUP_data %>%
  map(~ count(tibble(x = .x), x))
```

## Preprocess data

### IHC scores

Data wrangling - split multiple IHC scores into separate columns. Deliminators can be comma or semicolon

```{r}
CUP_data_wrangled <- CUP_data |>
  # split_to_one_row_per_measurement(column_name = `IHC scores`, measurement_name = "IHC") |>
  # Keep the original column before processing
  mutate(value_name = `IHC scores`) |>
  # replace "na" by NA
  mutate(value_name = ifelse(value_name == "na", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  separate_wider_delim(value_name,
    names = c("col_name", "value_name"),
    delim = " ",
    too_few = "align_start",
    too_many = "merge"
  ) |>
  # Fix problem that MMR can either be d or p instead of a number behind the name
  mutate(value_name = ifelse(col_name %in% c("pMMR", "dMMR"), col_name, value_name)) |>
  mutate(col_name = ifelse(col_name %in% c("pMMR", "dMMR"), "MMR", col_name)) |>
  mutate(col_name = ifelse(is.na(col_name), NA, paste0("IHC_", col_name))) |>
  df_wider_and_cleanup()

# Quality control, check that columns with all NA for IHC were indeed empty
CUP_data_wrangled |>
  filter(if_all(starts_with("IHC_"), is.na)) |>
  pull(CUPID)

# Look at the output
CUP_data_wrangled |>
  select(CUPID, starts_with("IHC")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Include a measure if IHCs were assessed.


### SNVs

Replace columns with only x by NA, because it combines "not assessed", "not detected" and "not relevant". 
And after that split mutations into a column per gene, where the value is the type of SNV. 

Note: what to do when two SNVs are found in one gene? I now put them in the same field separated by |

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant SNVs`) |>
  # Replace X by NA
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a SNV is the gene
  mutate(col_name = ifelse(is.na(value_name), NA,
    paste0("SNV_", str_extract(value_name, "^[^ ]+"))
  )) |>
  df_wider_and_cleanup()

# Look at the intermediate
CUP_data_wrangled |>
  select(CUPID, contains("SNV")) |>
  head(3)
```

# Create a list of all SNVs in this study

```{r}
all_SNVs <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant SNVs`) |>
  # Replace X by NA
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a SNV is the gene
  mutate(gene_symbol = ifelse(is.na(value_name), NA,
    str_extract(value_name, "^[^ ]+")
  )) |>
  select(gene_symbol, value_name) |>
  remove_missing() |>
  arrange(gene_symbol)

output_file <- file.path(file_location, "output", str_glue("all_SNVs_{cohort_name}.csv"))
print(paste("Saving to...", output_file))
write_csv(all_SNVs,
  file = output_file,
  na = ""
)
```

#### Include whether genes were assessed for SNV analysis

Options are: assessed (no SNVs identified), not assessed, or the name of the SNV per gene.

The code below:
1. Uses the Method column to decide which gene panels were used.
2. For each gene SNV, it replaces the "NA" with either assessed or not assessed.

```{r}
gene_panels <- read_tsv(file.path(file_location, "data/gene_metadata", gene_panels_tsv))
head(gene_panels)

panel_per_platform_lookup <- read_tsv(file.path(file_location, "data/gene_metadata", panel_per_platform_lookup_tsv))

source(here("code/R/functions/genes_assessed.R"))

# 1. Create a list of assessed SNVs per patient
genes_assessed_per_patient <- map(CUP_data_wrangled$Methods, get_list_of_assessed_genes_per_patient, mutation_type = "SNV")

# 2. Check for each gene symbol if it is assessed
CUP_data_wrangled <- CUP_data_wrangled |>
  mutate(patient_index = row_number()) |>
  rowwise() |>
  mutate(
    across(starts_with("SNV_"), \(x) set_assessment_value(x, patient_index = patient_index, col_name = cur_column()))
  ) |>
  select(-patient_index)
CUP_data_wrangled |>
  select(CUPID, Methods, contains("SNV")) |>
  head(3)
```

#### Quality control

```{r}
# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("SNV")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Relevant deletions

This can be not assessed, assessed (but nothing found or not relevant) or the gene
name itself.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant Deletions`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  mutate(col_name = ifelse(is.na(value_name), NA, paste0("Del_", value_name))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Del")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Relevant Amplifications

Similar syntax as deletions.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant Amplifications`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  mutate(col_name = ifelse(is.na(value_name), NA, paste0("Amp_", value_name))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Amp")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Overexpression/Signatures

This column can have genes and signatures. Use the same approach as in SNVs to
preserve information in the columns as values.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Overexpression/Signatures`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a signature is the gene/signature. Or "not assessed"
  mutate(col_name = ifelse(is.na(value_name), NA,
    ifelse(value_name == "not assessed", "Sig_not_assessed",
      paste0("Sig_", str_extract(value_name, "^[^ ]+"))
    )
  )) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Sig")) |>
  head(2)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```


## Sort per gene and transpose table

For now the sorting is alphabetical.

```{r}
new_colnames <- setdiff(colnames(CUP_data_wrangled), colnames(CUP_data))
new_colnames
CUP_data_transposed <- CUP_data_wrangled |>
  # Move the gene name first so that we can sort them
  rename_with(
    ~ sub("(.*?)_(.*)", "\\2_\\1", .x),
    .cols = all_of(new_colnames)
  )

# Sort columns
new_colnames_sorted <- setdiff(colnames(CUP_data_transposed), colnames(CUP_data)) |>
  sort()

CUP_data_transposed <- CUP_data_transposed |>
  select(all_of(colnames(CUP_data)), all_of(new_colnames_sorted)) |>
  # Transpose
  pivot_longer(-CUPID, names_to = "col", values_to = "value") |>
  mutate(CUPID = paste0("Patient ", CUPID)) |>
  pivot_wider(names_from = CUPID, values_from = value)


colnames(CUP_data_transposed)
head(CUP_data_transposed, 1)
```

## Save preprocessed data

```{r}
if (!dir.exists(file.path(file_location, "output"))) dir.create(file.path(file_location, "output"))
output_file <- file.path(file_location, "output", str_replace(input_excel, ".xlsx$", "_preprocessed.csv"))
print(paste("Saving to...", output_file))
write_csv(CUP_data_wrangled,
  file = output_file,
  na = ""
)

output_file <- file.path(file_location, "output", str_replace(input_excel, ".xlsx$", "_transposed.csv"))
print(paste("Saving to...", output_file))
write_csv(CUP_data_transposed,
  file = output_file,
  na = ""
)
```

### Save in the required format for 02_visualization.py

Mayb clinical_metadata would be a better name? Or something similar to CBioPortal?

```{r}
lookup <- c(
  # new = old
  "Patient_identifier" = "CUPID",
  "Sex"                = "Sex (M/F)",
  "Rec."               = "MMT recommended (y/n)",
  "Rec. nr."           = "MMT recommended (number; na/not applicable)",
  "Diagnosis adapted"  = "Diagnosis adapted (y/n)"
)
CUP_data_wrangled |>
  rename(any_of(lookup)) |>
  select(
    "Patient_identifier", "Histology", "Sex", "Rec.", "Rec. nr.",
    "Diagnosis adapted"
  ) |>
  write_csv(file = file.path(
    file_location, "output",
    str_glue("{study_date}_{cohort_name}_Entities.csv")
  ))
```
