---
title: "01_data_preparation_CUP"
date: "`r Sys.Date()`"
engine: knitr
format: 
  html:
    self-contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    df-print: kable
---

## Introduction

This draft script is used to analyze data in a different format than the previously
shared 01_data_preparation.ipynb notebook. Documentation will be updated.

## Preparation

```{r}
library(tidyverse)
library(here)
library(readxl)
```

Define functions


```{r}
df_wider_and_cleanup <- function(df){
  df <- df |>
    mutate(col_name = str_replace_all(col_name, " ", "_")) |>
    pivot_wider(
      names_from = col_name, 
      values_from = value_name,
      values_fill = NA
    )
  if(any(colnames(df) == "NA")){
    df <- select(df, -"NA")
  }
  return(df)
}
```

## Load data

```{r}
input_excel <- "Oncoprin_CUPxMTB_fakedata.xlsx"
CUP_data <- read_excel(here("data", input_excel))

dim(CUP_data)
head(CUP_data)
colnames(CUP_data)

nr_of_samples <- nrow(CUP_data)
```

Study the data structure

```{r}
CUP_data %>% 
  map(~ count(tibble(x = .x), x))
```

## Preprocess data

### IHC scores
Data wrangling - split multiple IHC scores into separate columns. Deliminators can be comma or semicolon

```{r}
CUP_data_wrangled <- CUP_data |>
  #split_to_one_row_per_measurement(column_name = `IHC scores`, measurement_name = "IHC") |>
  # Keep the original column before processing
  mutate(value_name = `IHC scores`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  separate_wider_delim(value_name, names = c("col_name", "value_name"), 
                       delim = " ", 
                       too_few = "align_start",
                       too_many = "merge") |>
  # Fix problem that MMR can either be d or p instead of a number behind the name
  mutate(value_name = ifelse(col_name %in% c("pMMR", "dMMR"), col_name, value_name)) |>
  mutate(col_name = ifelse(col_name %in% c("pMMR", "dMMR"), "MMR", col_name)) |>
  mutate(col_name = ifelse(is.na(col_name), NA, paste0("IHC_", col_name))) |>
  df_wider_and_cleanup()

# Quality control, check that columns with all NA for IHC were indeed empty
CUP_data_wrangled |>
  filter(if_all(starts_with("IHC_"), is.na)) |>
  pull(CUPID)

# Look at the output
CUP_data_wrangled |>
  select(CUPID, starts_with("IHC")) |> head()

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### SNVs

Replace columns with only x by NA, because it combines "not assessed", "not detected" and "not relevant". 
And after that split mutations into a column per gene, where the value is the type of SNV. 

Note: what to do when two SNVs are found in one gene?

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant SNVs`) |>
  # Replace X by NA
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |> 
  # Assume that the first word of a SNV is the gene
  mutate(col_name = ifelse(is.na(value_name), NA,
    paste0("SNV_", str_extract(value_name, "^[^ ]+")))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("SNV")) |> head()

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

## Relevant deletions

This can be not assessed, assessed (but nothing found or not relevant) or the gene
name itself.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant Deletions`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |> 
  mutate(col_name = ifelse(is.na(value_name), NA, paste0("Del_", value_name))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Del")) |> head()

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

## Relevant Amplifications

Similar syntax as deletions.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant Amplifications`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |> 
  mutate(col_name = ifelse(is.na(value_name), NA, paste0("Amp_", value_name))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Amp")) |> head()

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

## Overexpression/Signatures

This column can have genes and signatures. Use the same approach as in SNVs to
preserve information in the columns as values.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Overexpression/Signatures`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a signature is the gene/signature. Or "not assessed"
  mutate(col_name = ifelse(is.na(value_name), NA, 
           ifelse(value_name == "not assessed", "Sig_not_assessed",
    paste0("Sig_", str_extract(value_name, "^[^ ]+"))))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Sig")) |> head()

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```


## Save preprocessed data

```{r}
write_csv(CUP_data_wrangled,
          file = here("data", str_replace(input_excel, ".xlsx$", "_preprocessed.csv")),
          na = "")
```

  