---
title: "03_metadata_gene_lists"
date: "`r Sys.Date()`"
engine: knitr
format: 
  html:
    self-contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    df-print: kable
---

## Introduction

Create a table to show which genes were assessed in which molecular diagnostics assay.
Visualize this using an UpSet plot.

### Note on renaming old gene names

Some gene symbols were no longer used. The HGNC Multi Symbol Checker (https://www.genenames.org/tools/multi-symbol-checker/) and the HGNC approved gene lists with version 2026-04-07 (hgnc_complete_set_2026-04-07.tsv) were used to check for old gene symbols. The original gene symbol, change, and URL are listed in the Excel sheets when any changes were made.

## Preparation

```{r}
library(tidyverse)
library(here)
library(readxl)
library(UpSetR)
```

## Load metadata

Open the HGNC table with gene symbols and HGNC identifiers.

```{r}
file_location <- here()

gene_identifier_table <- read_tsv(file.path(
  file_location, "data/gene_metadata",
  "hgnc_complete_set_2026-04-07.tsv"
)) |>
  rename(gene_symbol = symbol)
```
Read all sheets in the metadata table, except the index sheet.

```{r}
input_metadata <- "metadata_panels_and_sequencing.xlsx"
(input_metadata <- file.path(file_location, "data/gene_metadata", input_metadata))

metadata_sheets <- setdiff(excel_sheets(input_metadata), "index")
metadata_sheets

metadata <- lapply(metadata_sheets, \(x) read_excel(path = input_metadata, sheet = x))
names(metadata) <- metadata_sheets
```

Combine all sheets with assessed genes into one table.

```{r}
# full join on gene symbols to get all gene symbols into a data frame
gene_metadata_merged <- metadata |>
  imap(\(sheet, column) sheet |>
    mutate("platform" = column) |>
    select(platform, gene_symbol)) |>
  bind_rows() |>
  mutate(gene_symbol2 = gene_symbol) |>
  pivot_wider(id_cols = gene_symbol, names_from = platform, values_from = gene_symbol2) |>
  mutate(across(everything(), \(x) replace_na(x, "not assessed"))) |>
  left_join(
    select(gene_identifier_table, hgnc_id, entrez_id, gene_symbol, ensembl_gene_id),
    by = "gene_symbol"
  )
head(gene_metadata_merged)
dim(gene_metadata_merged)
```

## Include all genes for MASTER and WGS

For the analysis with DKTK/MASTER or WGS, whole genome sequencing was used.

This means (we assume) that all genes were measured. We could either add a list of genes (e.g >20.000 genes), or only include the genes listed in other assays in tables, and allow "all assessed" in further analyses.

## Write to file

```{r}
write_tsv(gene_metadata_merged,
  (file.path(file_location, "data/gene_metadata", "metadata_panels_and_sequencing.tsv")),
  na = ""
)
# Also save the index table with lists-to-use
index_table <- read_excel(path = input_metadata, sheet = "index") |>
  select(Platform, `Gene list SNVs`, `Gene list Fusions`, `SNVs detected`, `Fusions detected`)

index_table |>
  write_tsv(
    (file.path(file_location, "data/gene_metadata", "gene_panels_and_platform_names.tsv")),
    na = ""
  )
```

## Synthetic metadata

Based on: https://www.statology.org/create-empty-data-frame-in-r/

```{r}
# create data frame with 0 rows and 5 columns
gene_metadata_synthetic <- data.frame(matrix(
  data = rep(c("A", "B", "C", "D"), ncol(gene_metadata_merged)),
  ncol = ncol(gene_metadata_merged), nrow = 4
))

# provide column names
colnames(gene_metadata_synthetic) <- colnames(gene_metadata_merged)


# Set some to not assessed
gene_metadata_synthetic[c(1, 2), "DHS-3501Z"] <- "not assessed"
gene_metadata_synthetic[c(1, 4), "Archer"] <- "not assessed"
gene_metadata_synthetic[c(2, 3), "Basic-NGS"] <- "not assessed"
gene_metadata_synthetic[c(1, 2), "Master"] <- "not assessed"

# Stuff
gene_metadata_synthetic["NCBI_Gene_ID"] <- c(1, 2, NA, 4)
gene_metadata_synthetic["Ensembl_gene_ID"] <- c("ENSG1", "ENSG2", NA, "ENSG4")
gene_metadata_synthetic

write_tsv(gene_metadata_synthetic,
  (file.path(file_location, "data/gene_metadata", "metadata_panels_and_sequencing_synthetic.tsv")),
  na = ""
)
```

## Check overlap between gene lists

First, study genes without a HGNC code. These are likely old gene names and should not be in the table.

```{r}
gene_metadata_merged |>
  filter(is.na(hgnc_id))
```

Next, prepare the data for visualization in an UpSet plot, where assessed genes have value 1 and not assessed genes value 0. 

Code based on: <https://upset.app/> and <https://github.com/hms-dbmi/UpSetR?tab=readme-ov-file>.

```{r}
data_for_upsetplot <- gene_metadata_merged |>
  select(gene_symbol, all_of(metadata_sheets)) |>
  rowwise() |>
  mutate(
    across(-gene_symbol, \(x) case_when(x == "not assessed" ~ as.integer(0),
      x == gene_symbol ~ as.integer(1),
      .default = NA
    ))
  ) |>
  column_to_rownames("gene_symbol")

if (any(is.na(data_for_upsetplot))) stop()

head(data_for_upsetplot)
```

Quality control: look at some values:

```{r}
table(data_for_upsetplot$Archer)
```

```{r}
index_table
```

```{r}
#| fig-height: 4.7 # 12 cm
#| fig-width: 6.3 # 16 cm
# SNVs
SNV_gene_lists <- index_table |>
  filter(`SNVs detected` == "yes") |>
  unique() |>
  pull(`Gene list SNVs`) |>
  unique()

upset_plot <- function(df){
  p <- upset(
    df, 
    nintersects = NA, 
    nsets = ncol(data_for_upsetplot) + 1,
    text.scale = 1.3,
    mb.ratio = c(0.65, 0.35)
    )
}
p_upset_SNV <- data_for_upsetplot |>
  select(intersect(colnames(data_for_upsetplot), SNV_gene_lists)) |>
  # Everything is measured in WGS and in MASTER, but that is >20.000 genes so only include overlap with other panels. Keep all rows with at least one 1, and set all for WGS to 1:
  filter_out(if_all(everything(), \(x) x == 0)) |>
  mutate("WGS-MASTER" = as.integer(1)) |>
  upset_plot()
p_upset_SNV

# Fusions
Fusion_gene_lists <- index_table |>
  filter(`Fusions detected` == "yes") |>
  unique() |>
  pull(`Gene list Fusions`)

p_upset_Fus <- data_for_upsetplot |>
  select(intersect(colnames(data_for_upsetplot), Fusion_gene_lists)) |>
  filter_out(if_all(everything(), \(x) x == 0)) |>
  mutate("WGS-MASTER" = as.integer(1)) |>
  upset_plot()
p_upset_Fus
```

#### Save figures

```{r}
# Save as png and svg
save_plot <- function(plot_obj, file_path, filename, verbose = TRUE){
  png_file <- file.path(file_path, str_glue("{filename}.png"))
  svg_file <- file.path(file_path, str_glue("{filename}.svg"))
  if(verbose){
    print(str_glue("Saving PNG to: {png_file}"))
    print(str_glue("Saving SVG to: {svg_file}"))
  }
  png(png_file, width = 16, height = 12, units = "cm", res = 300)
  show(plot_obj)
  dev.off()
  
  svg(svg_file, width = 16/2.54, height = 12/2.54)
  show(plot_obj)
  dev.off()
}
save_plot(p_upset_Fus, 
          file.path(file_location, "output/gene_metadata"), 
          "upset_plot_fusions")
save_plot(p_upset_SNV, 
          file.path(file_location, "output/gene_metadata"), 
          "upset_plot_SNVs")
```
