---
title: "03_metadata_gene_lists"
date: "`r Sys.Date()`"
engine: knitr
format: 
  html:
    self-contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    df-print: kable
---

## Introduction

Merge metadata lists with gene symbols

## Preparation

```{r}
library(tidyverse)
library(here)
library(readxl)
```

## Load metadata

```{r}
input_metadata <- "metadata_panels_and_sequencing.xlsx"
file_location <- here()
(input_metadata <- file.path(file_location, "data/gene_metadata", input_metadata))

gene_identifier_table <- read_csv(file.path(file_location, "data/gene_metadata",
                                   "20250116_gene_identifier_table_unique.csv"))
```

```{r}
metadata_sheets <- setdiff(excel_sheets(input_metadata), "index")
metadata_sheets

metadata <- lapply(metadata_sheets, \(x) read_excel(path = input_metadata, sheet = x))
names(metadata) <- metadata_sheets

# full join on gene symbols to get all gene symbols into a data frame
gene_metadata_merged <- metadata |>
  imap(\(sheet, column) sheet |> 
         mutate("platform" = column) |> 
         select(platform, gene_symbol)) |>
  bind_rows() |>
  mutate(gene_symbol2 = gene_symbol) |>
  pivot_wider(id_cols = gene_symbol, names_from = platform, values_from = gene_symbol2) |>
  mutate(across(everything(), \(x) replace_na(x, "not assessed"))) |>
  left_join(
    select(gene_identifier_table, NCBI_Gene_ID, gene_symbol, Ensembl_gene_ID, HGNC_ID),
    by = "gene_symbol"
  )
head(gene_metadata_merged)
dim(gene_metadata_merged)
```

## What to do with old gene names?

Use https://www.genenames.org/tools/multi-symbol-checker/ to get correct info for the missing genes and add it?

## Include all genes for MASTER WGS

Analysis within DKTK/MASTER whole genome sequencing + RNAseq
WGS: (whole genome sequencing)

This means (we assume that) that all genes were measured.

## Write to file

```{r}
write_tsv(gene_metadata_merged,
          (file.path(file_location, "data/gene_metadata", "metadata_panels_and_sequencing.tsv")),
          na = ""
          )
# Also save the index table with lists-to-use
read_excel(path = input_metadata, sheet = "index") |>
  select(Platform, Gene_list_to_use, `SNVs detected`, `Fusions detected`) |>
  write_tsv(
          (file.path(file_location, "data/gene_metadata", "gene_panels_and_platform_names.tsv")),
          na = ""
          )

```

## Synthetic metadata

Based on: https://www.statology.org/create-empty-data-frame-in-r/

```{r}
#create data frame with 0 rows and 5 columns
gene_metadata_synthetic <- data.frame(matrix(
  data = rep(c("A", "B", "C", "D"), ncol(gene_metadata_merged)),
  ncol = ncol(gene_metadata_merged), nrow = 4)
  )

#provide column names
colnames(gene_metadata_synthetic) <- colnames(gene_metadata_merged)


# Set some to not assessed
gene_metadata_synthetic[c(1,2),"DHS-3501Z"] <- "not assessed"
gene_metadata_synthetic[c(1,4), "Archer"] <- "not assessed"
gene_metadata_synthetic[c(2,3), "Basic-NGS"] <- "not assessed"
gene_metadata_synthetic[c(1,2), "Master"] <- "not assessed"

# Stuff
gene_metadata_synthetic["NCBI_Gene_ID"] <- c(1,2,NA,4)
gene_metadata_synthetic["Ensembl_gene_ID"] <- c("ENSG1","ENSG2",NA,"ENSG4")
gene_metadata_synthetic

write_tsv(gene_metadata_synthetic,
          (file.path(file_location, "data/gene_metadata", "metadata_panels_and_sequencing_synthetic.tsv")),
          na = ""
          )
```

