---
title: "01_data_preparation_CUP"
date: "`r Sys.Date()`"
engine: knitr
format: 
  html:
    self-contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    df-print: kable
---

## Introduction

This script processes a dataset where mutations and structural variants are written in one column, split by comma's, to a new format where each gene has its own column.

This can later be used for an OnkoPrint.

## Preparation

```{r}
library(tidyverse)
library(here)
library(readxl)
```

Define functions

```{r}
df_wider_and_cleanup <- function(df) {
  df <- df |>
    mutate(col_name = str_replace_all(col_name, " ", "_")) |>
    pivot_wider(
      names_from = col_name,
      values_from = value_name,
      values_fill = NA,
      values_fn = ~ paste(.x, collapse = "|")
    )
  if (any(colnames(df) == "NA")) {
    df <- select(df, -"NA")
  }
  if (any(colnames(df) == "SNV_NA")) {
    df <- select(df, -"SNV_NA")
  }
  return(df)
}
```

## Load data

```{r}
file_location <- here()
input_excel <- "Oncoprin_CUPxMTB_syntheticdata_V3.2.xlsx"

# Already list the table of gene panels used. This is needed later to check whether genes were assessed
gene_panels_tsv <- "metadata_panels_and_sequencing_synthetic.tsv"
panel_per_platform_lookup_tsv <- "gene_panels_and_platform_names.tsv"
study_date <- Sys.Date()
cohort_name <- "CUP_synthetic"

CUP_data <- read_excel(file.path(file_location, "data", input_excel))
dim(CUP_data)
head(CUP_data, 1)
colnames(CUP_data)

(nr_of_samples <- nrow(CUP_data))
```

Create an output directory using the cohort name and datestamp

```{r}
if (!dir.exists(file.path(file_location, "output"))) dir.create(file.path(file_location, "output"))

output_dir <- file.path(file_location, "output", str_c(Sys.Date(), cohort_name, sep = "_"))
if (!dir.exists(output_dir)) dir.create(output_dir)
```

Study the data structure

```{r}
CUP_data %>%
  map(~ count(tibble(x = .x), x))
```

## Preprocess data

### IHC scores

Data wrangling - split multiple IHC scores into separate columns. Deliminators can be comma or semicolon.

```{r}
CUP_data_wrangled <- CUP_data |>
  # split_to_one_row_per_measurement(column_name = `IHC scores`, measurement_name = "IHC") |>
  # Keep the original column before processing
  mutate(value_name = `IHC scores`) |>
  # replace "na" by NA
  mutate(value_name = ifelse(value_name == "na", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  separate_wider_delim(value_name,
    names = c("col_name", "value_name"),
    delim = " ",
    too_few = "align_start",
    too_many = "merge"
  ) |>
  # Fix problem that MMR can either be d or p instead of a number behind the name
  mutate(value_name = ifelse(col_name %in% c("pMMR", "dMMR"), col_name, value_name)) |>
  mutate(col_name = ifelse(col_name %in% c("pMMR", "dMMR"), "MMR", col_name)) |>
  mutate(col_name = ifelse(is.na(col_name), NA, paste0("IHC_", col_name))) |>
  df_wider_and_cleanup()

# Quality control, check that columns with all NA for IHC were indeed empty
CUP_data_wrangled |>
  filter(if_all(starts_with("IHC_"), is.na)) |>
  pull(CUPID)

# Look at the output
CUP_data_wrangled |>
  select(CUPID, starts_with("IHC")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Relevant SNVs

Replace columns with only x by NA, because it combines "not assessed", "not detected" and "not relevant". And after that split mutations into a column per gene, where the value is the type of SNV.

Note: what to do when two SNVs are found in one gene? I now put them in the same field separated by \|

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant SNVs`) |>
  # Replace X by NA
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a SNV is the gene
  mutate(col_name = ifelse(is.na(value_name), NA,
    paste0("SNV_", str_extract(str_trim(value_name), "^[^ ]+"))
  )) |>
  df_wider_and_cleanup()

# Look at the intermediate
CUP_data_wrangled |>
  select(CUPID, contains("SNV")) |>
  head(3)
```

## Create a list of all relevant SNVs in this study

```{r}
all_SNVs <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant SNVs`) |>
  # Replace X by NA
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a SNV is the gene
  mutate(gene_symbol = ifelse(is.na(value_name), NA,
    str_extract(str_trim(value_name), "^[^ ]+")
  )) |>
  select(gene_symbol, value_name) |>
  remove_missing() |>
  arrange(gene_symbol)

output_file <- file.path(output_dir, str_glue("all_SNVs_{cohort_name}.csv"))
print(paste("Saving to...", output_file))
write_csv(all_SNVs,
  file = output_file,
  na = ""
)
```

### Also include rows for non-relevant SNVs

For genes that were altered but not relevant: create a column if it was measured, but not in the column of relevant SNVs. Do not include the mutation itself, only "altered - not relevant".

```{r}
all_SNVs_CUP <- CUP_data_wrangled |>
  select(CUPID, `All SNVs`) |>
  rename(value_name = `All SNVs`) |>
  # Replace X by NA
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  mutate(value_name = ifelse(value_name == "not detected", NA, value_name)) |>
  mutate(value_name = ifelse(value_name == "not assessed", NA, value_name)) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a SNV is the gene
  mutate(col_name = ifelse(is.na(value_name), NA,
    paste0("SNV_", str_extract(str_trim(value_name), "^[^ ]+"))
  )) |>
  # Set all values to "altered - not relevant"
  mutate(value_name = ifelse(is.na(value_name), NA, "altered - not relevant")) |>
  distinct() |>
  df_wider_and_cleanup()

# Add only the columns for which no relevant SNV was measured. In the next section
# we add the "altered - not relevant" to SNV columns with relevant mutations.
# but that is a bit more complicated
columns_to_add <- setdiff(colnames(all_SNVs_CUP), colnames(CUP_data_wrangled))

CUP_data_wrangled <- CUP_data_wrangled |>
  left_join(
    select(all_SNVs_CUP, CUPID, all_of(columns_to_add)),
    by = "CUPID"
  )
```


#### Include whether genes were assessed for SNV analysis

Options are: assessed (no SNVs identified), not assessed, not relevant, or the name of the SNV per gene.

The code below:

1.  Uses the Method column to decide which gene panels were used.
2.  For each gene SNV, it replaces the "NA" with either assessed or not assessed.
3.  Checks if the gene is in the column "all SNVs", meaning that there was a mutation, but not a relevant one.

```{r}
gene_panels <- read_tsv(file.path(file_location, "data/gene_metadata", gene_panels_tsv))
head(gene_panels)

panel_per_platform_lookup <- read_tsv(file.path(file_location, "data/gene_metadata", panel_per_platform_lookup_tsv))

source(here("R/functions/genes_assessed.R"))

# 1. Create a list of assessed SNVs per patient
genes_assessed_per_patient <- map(CUP_data_wrangled$Methods,
  get_list_of_assessed_genes_per_patient,
  mutation_type = "SNV",
  panel_lookup = panel_per_platform_lookup
)

# 2. Create a list of altered SNVs per patient, including non-relevant ones
genes_altered_per_patient <- CUP_data_wrangled |>
  # Replace X by NA
  mutate(value_name = `All SNVs`) |>
  mutate(value_name = ifelse(value_name == "x", NA, value_name)) |>
  # Extract all genes
  mutate(
    altered_genes = map(
      str_split(value_name, ","), \(x) str_extract(str_trim(x), "^[^ ]+")
    )
  ) |>
  select(CUPID, altered_genes) |>
  deframe()

# 2. Check for each gene symbol if it is assessed and relevant
CUP_data_wrangled <- CUP_data_wrangled |>
  mutate(patient_index = row_number()) |>
  rowwise() |>
  mutate(
    across(starts_with("SNV_"), \(x) set_assessment_value(
      x,
      patient_index = patient_index,
      col_name = cur_column(),
      genes_assessed = genes_altered_per_patient,
      assessed_label = "altered - not relevant",
      not_assessed_label = NA
    ))
  ) |>
  mutate(
    across(starts_with("SNV_"), \(x) set_assessment_value(
      x,
      patient_index = patient_index,
      col_name = cur_column(),
      genes_assessed = genes_assessed_per_patient,
      assessed_label = "assessed",
      not_assessed_label = "not assessed"
    ))
  ) |>
  select(-patient_index)
CUP_data_wrangled |>
  select(CUPID, Methods, contains("SNV")) |>
  head(3)
```

#### Create a small report with the nr of assessed genes per patient

```{r}
stopifnot(length(genes_assessed_per_patient) == nrow(CUP_data_wrangled))
names(genes_assessed_per_patient) <- CUP_data_wrangled$CUPID

## Patients assessed per gene?
assessed_SNVs_per_patient <- get_patients_assessed_per_gene(genes_assessed_per_patient) |>
  rename(CUPID = ID) |>
  mutate(CUPID = paste0("Patient ", CUPID))
```

#### Quality control

```{r}
# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("SNV")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Relevant deletions

This can be 'not assessed', 'assessed' (but nothing found or not relevant) or the gene name itself.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant Deletions`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  mutate(col_name = ifelse(is.na(value_name), NA, paste0("Del_", value_name))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Del")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Relevant Amplifications

Similar syntax as deletions.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Relevant Amplifications`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  mutate(col_name = ifelse(is.na(value_name), NA, paste0("Amp_", value_name))) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Amp")) |>
  head(3)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

### Overexpression/Signatures

This column can have genes and signatures. Use the same approach as in SNVs to preserve information in the columns as values.

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  # Keep the original column before processing
  mutate(value_name = `Overexpression/Signatures`) |>
  separate_longer_delim(value_name, delim = regex("[;,] ")) |>
  # Assume that the first word of a signature is the gene/signature. Or "not assessed"
  mutate(col_name = case_when(
    is.na(value_name) ~ NA,
    value_name == "not assessed" ~ "Sig_not_assessed",
    .default = paste0("Sig_", str_extract(value_name, "^[^ ]+"))
  )) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Sig")) |>
  head(2)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

## Fusions

Replace "na" with "not assessed"

```{r}
CUP_data_wrangled <- CUP_data_wrangled |>
  mutate(`Fusions detected` = str_replace(`Fusions detected`, "^na$", "not assessed")) |>
  # Keep the original column before processing
  mutate(value_name = `Fusions detected`) |>
  separate_longer_delim(value_name, delim = regex("[,] ")) |>
  # A fusion is two genes/components split by two colons, e.g. A::B. By adding an extra split,
  # it should create two rows, one for the first gene fusion and one for the second half.
  # If there is additional info, e.g. the chromosome, we remove it here to only keep gene names
  # This is everything after a space after the gene fusions
  mutate(fusion_component = value_name) |>
  mutate(fusion_component = str_remove(fusion_component, " .*")) |>
  separate_longer_delim(fusion_component, delim = regex("::")) |>
  # A fusion is two genes/components split by two colons, e.g. A::B
  mutate(col_name = case_when(
    is.na(value_name) ~ NA,
    value_name == "not assessed" ~ "Fus_not_assessed",
    .default = paste0("Fus_", fusion_component)
  )) |>
  select(-fusion_component) |>
  df_wider_and_cleanup()

# Look at the output
CUP_data_wrangled |>
  select(CUPID, contains("Fus")) |>
  head(2)

# Quality control: rows should not be added or deleted
stopifnot(nrow(CUP_data_wrangled) == nr_of_samples)
```

#### Include whether genes were assessed for fusion analysis

Options are: 'assessed' (no fusions identified), 'not assessed', or the name of the fusion per gene.

The code below:

1.  Uses the Method column to decide which gene panels were used.
2.  For each gene fusion, it replaces the "NA" with either assessed or not assessed.

```{r}
# 1. Create a list of assessed fusions per patient
genes_assessed_per_patient <- map(CUP_data_wrangled$Methods,
  get_list_of_assessed_genes_per_patient,
  mutation_type = "Fusion",
  panel_lookup = panel_per_platform_lookup
)

# 2. Check for each gene symbol if it is assessed
CUP_data_wrangled <- CUP_data_wrangled |>
  mutate(patient_index = row_number()) |>
  rowwise() |>
  mutate(
    across(starts_with("Fus_"), \(x) set_assessment_value(x, patient_index = patient_index, col_name = cur_column(), genes_assessed = genes_assessed_per_patient))
  ) |>
  select(-patient_index, -Fus_assessed, -Fus_not_assessed)
CUP_data_wrangled |>
  select(CUPID, Methods, contains("Fus")) |>
  head(3)
```

#### Create a small report with the nr of assessed genes per patient

```{r}
stopifnot(length(genes_assessed_per_patient) == nrow(CUP_data_wrangled))
names(genes_assessed_per_patient) <- CUP_data_wrangled$CUPID

## Patients assessed per gene?
assessed_fusions_per_patient <- get_patients_assessed_per_gene(genes_assessed_per_patient) |>
  rename(CUPID = ID) |>
  mutate(CUPID = paste0("Patient ", CUPID))
```

## Sort per gene and transpose table

For now the sorting is alphabetical.

```{r}
sort_and_transpose_table <- function(CUP_data_wrangled) {
  new_colnames <- setdiff(colnames(CUP_data_wrangled), colnames(CUP_data))
  print(new_colnames)
  CUP_data_transposed <- CUP_data_wrangled |>
    # Move the gene name first so that we can sort them
    rename_with(
      ~ sub("(.*?)_(.*)", "\\2_\\1", .x),
      .cols = all_of(new_colnames)
    )

  # Sort columns
  new_colnames_sorted <- setdiff(colnames(CUP_data_transposed), colnames(CUP_data)) |>
    sort()

  CUP_data_transposed <- CUP_data_transposed |>
    select(all_of(colnames(CUP_data)), all_of(new_colnames_sorted)) |>
    # Transpose
    pivot_longer(-CUPID, names_to = "col", values_to = "value") |>
    mutate(CUPID = paste0("Patient ", CUPID)) |>
    pivot_wider(names_from = CUPID, values_from = value)


  print(colnames(CUP_data_transposed))
  print(head(CUP_data_transposed, 1))
  return(CUP_data_transposed)
}
CUP_data_transposed <- sort_and_transpose_table(CUP_data_wrangled)
```

## Save preprocessed data

Save files

```{r}
output_file <- file.path(output_dir, str_replace(input_excel, ".xlsx$", "_preprocessed.csv"))
print(paste("Saving to...", output_file))
write_csv(CUP_data_wrangled,
  file = output_file,
  na = ""
)

output_file <- file.path(output_dir, str_replace(input_excel, ".xlsx$", "_transposed.csv"))
print(paste("Saving to...", output_file))
write_csv(CUP_data_transposed,
  file = output_file,
  na = ""
)
```

## Create a separate file without exact mutations, only with affected genes

For KRAS and BRAF, we will use a subcategory:

For KRAS: highlight G12C mutations (see for example https://doi.org/10.1111/joim.13057 and https://doi.org/10.1038/s41392-025-02473-8)

For BRAF, highlight class 1 and 3 (see for example https://doi.org/10.1158/1078-0432.CCR-19-0311) Definition:

-   class 1–V600E
-   class 2–codons 597/601
-   class 3–codons 594/596

```{r}
define_BRAF_class <- function(BRAF_mut) {
  if (BRAF_mut == "not assessed") {
    return("not assessed")
  }
  if (BRAF_mut == "altered - not relevant") {
    return("altered - not relevant")
  }
  if (BRAF_mut == "assessed") {
    return("assessed")
  }

  # class 1–V600E
  if (BRAF_mut == "BRAF V600E") {
    return("BRAF class I")
  }
  BRAF_codon <- str_extract(BRAF_mut, "\\d+") |>
    as.integer()
  # class 2–codons 597/601
  if (BRAF_codon >= 597 & BRAF_codon <= 601) {
    return("BRAF class II")
  }
  # class 3–codons 594/596
  if (BRAF_codon >= 594 & BRAF_codon <= 596) {
    return("BRAF class III")
  }
  return("BRAF other")
}

define_KRAS_class <- function(KRAS_mut) {
  case_when(
    KRAS_mut == "not assessed" ~ "not assessed",
    KRAS_mut == "assessed" ~ "assessed",
    KRAS_mut == "altered - not relevant" ~ "altered - not relevant",
    KRAS_mut == "KRAS pG12C" ~ "KRAS p.G12C",
    KRAS_mut == "KRAS p.G12C" ~ "KRAS p.G12C",
    .default = "KRAS non-G12C"
  )
}
```


```{r}
CUP_data_wrangled_pergene <- CUP_data_wrangled |>
  # Replace specific mutations with the gene name
  # Assume that the first word of a SNV is the gene
  # Except to show the difference between KRAS G12C and other mutations
  # Or BRAF class I and III
  rowwise() |>
  mutate(
    across(any_of("SNV_BRAF"), define_BRAF_class),
    across(any_of("SNV_KRAS"), define_KRAS_class)
  ) |>
  mutate(across(starts_with("SNV_"), \(x) case_when(
    # KRAS
    str_detect(x, "KRAS") ~ x,
    # BRAF
    str_detect(x, "BRAF") ~ x,
    # Others SNVs
    x == "assessed" ~ "assessed",
    x == "not assessed" ~ "not assessed",
    x == "altered - not relevant" ~ "altered - not relevant",
    .default = str_extract(x, "^[^ ]+")
  ))) |>
  # Replace specific fusions with the gene name
  # For this, use the column name that I defined before
  mutate(across(starts_with("Fus_"), \(x) case_when(
    x == "assessed" ~ "assessed",
    x == "not assessed" ~ "not assessed",
    .default = str_remove(cur_column(), "^Fus_")
  ))) |>
  # Fill NA in the IHC with "assessed" or "not assessed".
  # Assumption: if there is no score, the specific IHC was not assessed.
  mutate(across(starts_with("IHC_"), \(x) ifelse(is.na(x), "not assessed", x))) |>
  # Clean up deletions and amplifications
  select(-Del_not_assessed, -Del_assessed) |>
  mutate(across(starts_with("Del_"), \(x) case_when(
    is.na(x) & `Relevant Deletions` == "not assessed" ~ "not assessed",
    is.na(x) ~ "assessed",
    .default = x
  ))) |>
  select(-Amp_not_assessed, -Amp_assessed) |>
  mutate(across(starts_with("Amp_"), \(x) case_when(
    is.na(x) & `Relevant Amplifications` == "not assessed" ~ "not assessed",
    is.na(x) ~ "assessed",
    .default = x
  ))) |>
  # Fill NA in the signatures with assessed or not assessed - if MASTER was used
  select(-Sig_not_assessed) |>
  mutate(across(starts_with("Sig_"), \(x) case_when(
    is.na(x) & str_detect(Methods, "MASTER") ~ "assessed",
    is.na(x) ~ "not assessed",
    .default = x
  )))
```

## Sort per gene and transpose table

For now the sorting is alphabetical.

```{r}
CUP_data_transposed_pergene <- sort_and_transpose_table(CUP_data_wrangled_pergene)
```

## Save preprocessed data

```{r}
output_file <- file.path(output_dir, str_replace(input_excel, ".xlsx$", "_preprocessed_pergene.csv"))
print(paste("Saving to...", output_file))
write_csv(CUP_data_wrangled_pergene,
  file = output_file,
  na = ""
)

output_file <- file.path(output_dir, str_replace(input_excel, ".xlsx$", "_transposed_pergene.csv"))
print(paste("Saving to...", output_file))
write_csv(CUP_data_transposed_pergene,
  file = output_file,
  na = ""
)

### Save the tables whether genes were assessed for SNV and fusion analysis
output_file <- file.path(
  output_dir,
  str_glue("assessed_SNVs_per_patient_{cohort_name}.csv")
)
print(paste("Saving to...", output_file))
write_csv(assessed_SNVs_per_patient,
  file = output_file,
  na = ""
)
output_file <- file.path(
  output_dir,
  str_glue("assessed_fusions_per_patient_{cohort_name}.csv")
)
print(paste("Saving to...", output_file))
write_csv(assessed_fusions_per_patient,
  file = output_file,
  na = ""
)
```
