---
title: "03_metadata_gene_lists"
date: "`r Sys.Date()`"
engine: knitr
format: 
  html:
    self-contained: true
    toc: true
    toc_depth: 3
    toc_float: true
    df-print: kable
---

## Introduction

Create a table to show which genes were assessed in which molecular diagnostics assay.
Visualize this using an UpSet plot.

### Note on renaming old gene names

Some gene symbols were no longer used. The HGNC Multi Symbol Checker (https://www.genenames.org/tools/multi-symbol-checker/) and the HGNC approved gene lists with version 2026-04-07 (hgnc_complete_set_2026-04-07.tsv) were used to check for old gene symbols. The original gene symbol, change, and URL are listed in the Excel sheets when any changes were made.

## Preparation

```{r}
library(tidyverse)
library(here)
library(readxl)
library(UpSetR)
```

## Load metadata

Open the HGNC table with gene symbols and HGNC identifiers.

```{r}
file_location <- here()

gene_identifier_table <- read_tsv(file.path(
  file_location, "data/gene_metadata",
  "hgnc_complete_set_2026-04-07.tsv"
)) |>
  rename(gene_symbol = symbol)
```
Read all sheets in the metadata table, except the index sheet.

```{r}
input_metadata <- "metadata_panels_and_sequencing_v1.0.xlsx"
(input_metadata <- file.path(file_location, "data/gene_metadata", input_metadata))

metadata_sheets <- setdiff(excel_sheets(input_metadata), "index")
metadata_sheets

metadata <- lapply(metadata_sheets, \(x) read_excel(path = input_metadata, sheet = x))
names(metadata) <- metadata_sheets
```

Combine all sheets with assessed genes into one table.

```{r}
# full join on gene symbols to get all gene symbols into a data frame
gene_metadata_merged <- metadata |>
  imap(\(sheet, column) sheet |>
    mutate("platform" = column) |>
    select(platform, gene_symbol)) |>
  bind_rows() |>
  mutate(gene_symbol2 = gene_symbol) |>
  pivot_wider(id_cols = gene_symbol, names_from = platform, values_from = gene_symbol2) |>
  mutate(across(everything(), \(x) replace_na(x, "not assessed"))) |>
  left_join(
    select(gene_identifier_table, hgnc_id, entrez_id, gene_symbol, ensembl_gene_id),
    by = "gene_symbol"
  )
head(gene_metadata_merged)
dim(gene_metadata_merged)
```

## Include all genes for MASTER and WGS

For the analysis with DKTK/MASTER or WGS, whole genome sequencing was used.

This means (we assume) that all genes were measured. We could either add a list of genes (e.g >20.000 genes), or only include the genes listed in other assays in tables, and allow "all assessed" in further analyses.

We have decided to keep this information in Figure captions and not include WGS/Master in the figure.

## Write to file

```{r}
write_tsv(gene_metadata_merged,
  (file.path(file_location, "data/gene_metadata", "metadata_panels_and_sequencing.tsv")),
  na = ""
)
# Also save the index table with lists-to-use
index_table <- read_excel(path = input_metadata, sheet = "index") |>
  select(Platform, `Gene list SNVs`, `Gene list Fusions`, `SNVs detected`, `Fusions detected`)

index_table |>
  write_tsv(
    (file.path(file_location, "data/gene_metadata", "gene_panels_and_platform_names.tsv")),
    na = ""
  )
```

## Synthetic metadata

Based on: https://www.statology.org/create-empty-data-frame-in-r/

```{r}
# create data frame with 0 rows and 5 columns
gene_metadata_synthetic <- data.frame(matrix(
  data = rep(c("A", "B", "C", "D"), ncol(gene_metadata_merged)),
  ncol = ncol(gene_metadata_merged), nrow = 4
))

# provide column names
colnames(gene_metadata_synthetic) <- colnames(gene_metadata_merged)


# Set some to not assessed
gene_metadata_synthetic[c(1, 2), "DHS-3501Z"] <- "not assessed"
gene_metadata_synthetic[c(1, 4), "Archer"] <- "not assessed"
gene_metadata_synthetic[c(2, 3), "Basic-NGS"] <- "not assessed"
gene_metadata_synthetic[c(1, 2), "Master"] <- "not assessed"

# Stuff
gene_metadata_synthetic["NCBI_Gene_ID"] <- c(1, 2, NA, 4)
gene_metadata_synthetic["Ensembl_gene_ID"] <- c("ENSG1", "ENSG2", NA, "ENSG4")
gene_metadata_synthetic

write_tsv(gene_metadata_synthetic,
  (file.path(file_location, "data/gene_metadata", "metadata_panels_and_sequencing_synthetic.tsv")),
  na = ""
)
```

## Check overlap between gene lists

First, study genes without a HGNC code. These are likely old gene names and should not be in the table.

```{r}
gene_metadata_merged |>
  filter(is.na(hgnc_id))
```

### UpSet Plot visualization

Next, prepare the data for visualization in an UpSet plot, where assessed genes have value 1 and not assessed genes value 0. 

Code based on: <https://upset.app/> and <https://github.com/hms-dbmi/UpSetR?tab=readme-ov-file>.

```{r}
data_for_upsetplot <- gene_metadata_merged |>
  select(gene_symbol, all_of(metadata_sheets)) |>
  rowwise() |>
  mutate(
    across(-gene_symbol, \(x) case_when(x == "not assessed" ~ as.integer(0),
      x == gene_symbol ~ as.integer(1),
      .default = NA
    ))
  ) |>
  column_to_rownames("gene_symbol")

if (any(is.na(data_for_upsetplot))) stop()

head(data_for_upsetplot)
```

Quality control: look at some values:

```{r}
table(data_for_upsetplot$Archer)
```

```{r}
index_table
```

```{r}
#| fig-height: 4.7 # 12 cm
#| fig-width: 6.3 # 16 cm
# SNVs
SNV_gene_lists <- index_table |>
  filter(`SNVs detected` == "yes") |>
  unique() |>
  pull(`Gene list SNVs`) |>
  unique()

upset_plot <- function(df){
  p <- upset(
    df, 
    nintersects = NA, 
    nsets = ncol(data_for_upsetplot) + 1,
    text.scale = 1.3,
    mb.ratio = c(0.65, 0.35)
    )
}

p_upset_SNV <- data_for_upsetplot |>
  select(intersect(colnames(data_for_upsetplot), SNV_gene_lists)) |>
  filter_out(if_all(everything(), \(x) x == 0)) |>
  upset_plot()
p_upset_SNV

# Fusions
Fusion_gene_lists <- index_table |>
  filter(`Fusions detected` == "yes") |>
  unique() |>
  pull(`Gene list Fusions`)

p_upset_Fus <- data_for_upsetplot |>
  select(intersect(colnames(data_for_upsetplot), Fusion_gene_lists)) |>
  filter_out(if_all(everything(), \(x) x == 0)) |>
  upset_plot()
p_upset_Fus
```

### Euler Plot visualization

Using the eulerr R package (https://jolars.github.io/eulerr/)

```{r}
library(eulerr)

data_for_eulerplot <- data_for_upsetplot |>
  select(intersect(colnames(data_for_upsetplot), SNV_gene_lists)) |>
  filter_out(if_all(everything(), \(x) x == 0))
```


The problem is that the 7 unique genes for DHS-3501Z are now missing. TODO discuss with Thomas what we want to keep.

```{r}
#| fig-height: 4.7 # 12 cm
#| fig-width: 6.3 # 16 cm

(p_euler_SNV <- plot(
  euler(data_for_eulerplot),
  edges = TRUE,
  labels = TRUE,
  quantities = list(cex = 0.7)
)
)

(p_euler_SNV_bw <- plot(
  euler(data_for_eulerplot),
  fills = FALSE,
  edges = TRUE,
  labels = TRUE,
  quantities = list(cex = 0.7)
)
)
```

```{r}
data_for_eulerplot |> colnames()
data_for_eulerplot |> 
  select("OFA", "OFA plus", "DHS-3501Z", "CGP") |>
  euler() |>
  plot(quantities = TRUE)
```

### Bar Plot visualization

Show overlap between techniques and genes in each list

```{r}
# Order from large to small
sorted_platform_names <- colSums(data_for_eulerplot) |> 
  sort(decreasing = TRUE) |> 
  names()

# Count overlap between techniques
df_plot <- data_for_eulerplot |>
  rownames_to_column("gene_symbol") |>
  pivot_longer(
    -gene_symbol,
    names_to = "platform",
    values_to = "assessed"
  ) |>
  group_by(gene_symbol) |>
  mutate(n_platforms_assessed = sum(assessed == 1, na.rm = TRUE)) |>
  ungroup() |>
  filter(assessed == 1) |>
  count(platform, n_platforms_assessed)

# Plot
p_bar <- df_plot |>
  mutate(n_platforms_assessed = factor(n_platforms_assessed),
         platform = factor(platform, levels = sorted_platform_names)) |>
  ggplot(aes(x = platform, y = n, fill = n_platforms_assessed)) +
  labs(x = "", y = "Number of assessed genes", fill = "Detected in nr. of platforms") +
  theme_classic() +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_line(color = "grey90")
) +
  guides(fill = guide_legend(nrow = 1))

(p_bar1 <- p_bar + 
  geom_col(color = "grey20") +
  scale_fill_grey(start = 1, end = 0)
)
(p_bar2 <- p_bar + 
  geom_col(color = "grey20") +
  scale_fill_grey(start = 1, end = 0) +
  geom_text(
    aes(label = ifelse(n > 20, n, "")),
    position = position_stack(vjust = 0.5),
    size = 3
  )
)
(p_bar3 <- p_bar + 
  geom_col() +
  scale_fill_grey(start = 0.9, end = 0)
)
(p_bar4 <- p_bar + 
  geom_col() +
  scale_fill_viridis_d(direction = -1)
)
```

### Save figures

```{r}
# Save as png and svg
save_plot <- function(plot_obj, file_path, filename, verbose = TRUE,
                      width = 16, height = 12
                      ){
  png_file <- file.path(file_path, str_glue("{filename}.png"))
  svg_file <- file.path(file_path, str_glue("{filename}.svg"))
  if(verbose){
    print(str_glue("Saving PNG to: {png_file}"))
    print(str_glue("Saving SVG to: {svg_file}"))
  }
  png(png_file, width = width, height = height, units = "cm", res = 300)
  show(plot_obj)
  dev.off()
  
  svg(svg_file, width = width/2.54, height = height/2.54)
  show(plot_obj)
  dev.off()
}
# UpSet plots
save_plot(p_upset_Fus, 
          file.path(file_location, "output/gene_metadata"), 
          "upset_plot_fusions")
save_plot(p_upset_SNV, 
          file.path(file_location, "output/gene_metadata"), 
          "upset_plot_SNVs")

# Euler plots
save_plot(p_euler_SNV, 
          file.path(file_location, "output/gene_metadata"), 
          "euler_plot_SNVs")
save_plot(p_euler_SNV_bw, 
          file.path(file_location, "output/gene_metadata"), 
          "euler_plot_SNVs_bw")

# Bar plots
save_plot(p_bar1, 
          file.path(file_location, "output/gene_metadata"), 
          "bar_plot_SNVs_option1")
save_plot(p_bar2, 
          file.path(file_location, "output/gene_metadata"), 
          "bar_plot_SNVs_option2")
save_plot(p_bar3, 
          file.path(file_location, "output/gene_metadata"), 
          "bar_plot_SNVs_option3")
save_plot(p_bar4, 
          file.path(file_location, "output/gene_metadata"), 
          "bar_plot_SNVs_option4")
```
